Sentinel lead pipe racks quantify orthophosphate's dose-response

### 2023-05-24 Rmarkdown/paper.Rmd
updates for journal submission

---
title: "Sentinel pipe racks quantify orthophosphate's effect on lead release into drinking water"
author: "Benjamin F. Trueman^§,^\\*, Wendy H. Krkošek^†^, and Graham A. Gagnon^§^"
date: "^§^Centre for Water Resources Studies, Department of Civil & Resource Engineering, Dalhousie University, 1360 Barrington St., Halifax, Nova Scotia, Canada B3H 4R2  \n&nbsp;  \n^†^Halifax Water, 450 Cowie Hill Rd., Halifax, Nova Scotia, Canada,  B3P 2V3  \n&nbsp;   \n&nbsp;*Corresponding author  \n&nbsp;E-mail: benjamin.trueman@dal.ca    \n&nbsp;Tel:902.494.6070  \n&nbsp;Fax: 902.494.3105"
output:
  word_document:
    reference_docx: StylesTemplate.docx
bibliography: references.bib
csl: environmental-science-and-technology.csl
---

@@

Orthophosphate is used widely to control lead release from plumbing into tap water. Its effect can be difficult to quantify, though, since tap water lead concentrations are site-specific. Sentinel homes with lead service lines are ideal for evaluating orthophosphate corrosion control programs, but best practices dictate the removal of lead service lines once they are identified. Sentinel homes, then, often have too short a useful life to be used effectively. Here we explore an alternative: sentinel pipe racks constructed with recovered lead pipes and supplied with water directly from the distribution system. We also propose a strategy for analyzing pipe rack data based on the generalized additive model, which approximates time series as a sum of smooth functions. In this study, geometric mean lead release from pipe racks exhibited a pronounced dose response, falling by `r round(abs(pcent_decrease$y))`% (95% credible interval: `r paste(suppressWarnings(sort(round(abs(with(pcent_decrease, c(ymin, ymax)))))), collapse = "--")`%) after an increase from 1 to 2 mg PO~4~ L^-1^, and then climbing by `r round(abs(pcent_increase$y))`% (95% credible interval: `r paste(suppressWarnings(sort(round(abs(with(pcent_increase, c(ymin, ymax)))))), collapse = "--")`%) after a decrease to 1.5 mg PO~4~ L^-1^. Data from the sentinel homes were largely consistent with those from pipe racks: geometric mean lead levels at the high orthophosphate dose (2 mg L^-1^) were `r d$d_50`% of those at the low dose (1 mg L^-1^, 95% credible interval: `r d$d_2.5`--`r d$d_97.5`%). Our results demonstrate sentinel pipe racks as a viable alternative to at-the-tap sampling for non-regulatory corrosion control monitoring. They also provide a fully Bayesian framework for quantifying orthophosphate's effect on lead release that is well-suited to incorporating information from multiple sources.

@@

@@

It can be difficult, though, to estimate orthophosphate's effect on lead in drinking water since lead concentrations are determined by site-specific plumbing characteristics. And while lead solubility modeling can be informative, it fails to account for the complex mineralogy of lead corrosion scale or the generation of particles [@tully_water_2019; @trueman_characterizing_2019]. A decrease in tap water lead sampled at sentinel homes over time is the most reliable metric of orthophosphate's success, and homes supplied by lead service lines represent the population most at-risk [@healthcanada2022]. But to protect the inhabitants' health, lead pipe is often replaced once identified. Sentinel homes, then, may have too short a life to be useful in monitoring plumbosolvency changes.

@@

where $T$ denotes the Student t-distribution with time-varying mean $\mu_t$, standard deviation $\sigma$, and degrees-of-freedom parameter $\nu$. The mean is modeled as the sum of smooth functions of time $f_j(t)$. The full model (Zones 1 and 2) included a pipe-specific intercept $\alpha_{pipe_i}$ and centered smooth terms, whereas the Zone 1 model included non-centered series-specific smooths and a global intercept ($\bar{\alpha}$ in place of $\alpha_{pipe_i}$ in equation $(1)$). The matrices $Z_j$ and $X_j$ represent the penalized and unpenalized basis functions comprising each of the $f_j(t)$, and $b_j$ and $\beta_j$ represent the penalized and unpenalized GAM coefficients. $Gamma$ and $N$ denote the gamma and normal distributions.

>>>>

---
title: "Sentinel lead pipe racks quantify orthophosphate's dose-response"
author: "Benjamin F. Trueman^§,^\\*, Wendy H. Krkošek^†^, and Graham A. Gagnon^§^"
date: "^§^Centre for Water Resources Studies, Department of Civil & Resource Engineering, Dalhousie University, 1360 Barrington St., Halifax, Nova Scotia, Canada B3H 4R2  \n&nbsp;  \n^†^Halifax Water, 450 Cowie Hill Rd., Halifax, Nova Scotia, Canada,  B3P 2V3  \n&nbsp;   \n&nbsp;*Corresponding author  \n&nbsp;E-mail: benjamin.trueman@dal.ca    \n&nbsp;Tel:902.494.6070  \n&nbsp;Fax: 902.494.3105"
output:
  word_document:
    reference_docx: StylesTemplate.docx
bibliography: references.bib
csl: environmental-science-and-technology.csl
---

@@

Orthophosphate is used to minimize lead contamination of tap water, but its benefits are difficult to quantify since lead concentrations are plumbing-dependent. Homes serviced by lead pipe are ideal for monitoring orthophosphate treatment, but best practices dictate the removal of lead once identified, which complicates sampling plans. Here we explore an alternative: recovered lead pipe racks supplied with distributed drinking water at various locations within a water system. We also propose a strategy for analyzing the data based on the generalized additive model, which approximates time series as a sum of smooth functions. In this study, geometric mean lead release from pipe racks exhibited a pronounced dose-response, falling by `r round(abs(pcent_decrease$y))`% after an increase from 1 to 2 mg PO~4~ L^-1^, and then climbing by `r round(abs(pcent_increase$y))`% after a decrease to 1.5 mg PO~4~ L^-1^. Data from nine sentinel homes were consistent with those from pipe racks: geometric mean lead at the high orthophosphate dose was `r d$d_50`% of that at the low dose. Our results demonstrate sentinel pipe racks as a viable alternative to at-the-tap sampling for non-regulatory corrosion control monitoring. They also provide a Bayesian framework for quantifying orthophosphate's effect on lead release that can incorporate information from multiple sources.

@@

# Synopsis

Sentinel pipe racks can be used to quantify the impact of planned and unplanned water quality changes on lead release.

# Graphical abstract

@@

It can be difficult, though, to estimate orthophosphate's effect on lead in drinking water since lead concentrations are determined by site-specific plumbing characteristics [@redmon_lead_2022]. And while modeling can be informative, it generally fails to account for the complex mineralogy of lead corrosion scale or---with a few notable exceptions [@abokifa_modeling_2017]---the generation of particles [@tully_water_2019; @trueman_characterizing_2019]. A decrease in tap water lead sampled at sentinel homes over time is the most reliable metric of orthophosphate's success, and homes supplied by lead service lines represent the population most at-risk [@healthcanada2022]. But to protect the inhabitants' health, lead pipe is often replaced once identified. Sentinel homes, then, may have too short a life to be useful in monitoring plumbosolvency changes.

@@

where $T$ denotes the Student t-distribution with time-varying mean $\mu_t$, standard deviation $\sigma$, and degrees-of-freedom parameter $\nu$. The mean is modeled as the sum of smooth functions of time $f_j(t)$. The full model (Zones 1 and 2) included a pipe-specific intercept $\alpha_{pipe_i}$ and centered smooth terms, whereas the Zone 1 model included non-centered series-specific smooths and a global intercept ($\bar{\alpha}$ in place of $\alpha_{pipe_i}$ in equation $(1)$). The matrices $Z_j$ and $X_j$ represent the penalized and unpenalized basis functions comprising each of the $f_j(t)$, and $b_j$ and $\beta_j$ represent the penalized and unpenalized GAM coefficients. The parameter $\phi$ is the first-order autoregressive coefficient, and $s$ represents the spacing in time between consecutive observations. $Gamma$ and $N$ denote the gamma and normal distributions.
